---
Title:  "Kia and Hyundai Theft Analysis"        
Author: "Brian Reppeto"                                                 
Date: "1/15/2025"  
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Introduction

"This report is an analysis of Kia and Hyundai vehicle thefts. The purpose is to identify patterns, trends, and actionable insights for policymakers and law enforcement agencies."

"The visualizations provide: - Proportions of theft types. - Monthly and yearly distributions. - Geospatial insights. - Key trends over time."

# Methodology

"The data was analyzed using R and visualized with the `ggplot2` package. Below are the results of the analysis."

# Load libraries

```{r setup, include=FALSE}
library(ggplot2) 
library(readr) 
library(openxlsx) 
library(tidyr) 
library(dplyr) 
library(treemap)
library(readxl)
library(maps) 
library(mapdata)
knitr::opts_chunk$set(echo = TRUE)
```

# Set the working directory

```{r work, include=TRUE}
setwd("C:/Users/brepp/BER Data Science/DSC640/Week5&6")
```

# Load CSV files

```{r load, include=TRUE}
car_map <- read_csv("carTheftsMap.csv") 
kiahyundaidata <- read_csv("KiaHyundaiMilwaukeeData.csv") 
kiahyundaithefts <- read_csv("kiaHyundaiThefts.csv") 
vicedata <- read_xlsx("Motherboard VICE News Kia Hyundai Theft Data.xlsx")
```

# Combine the datasets

```{r combine, include=TRUE}
combined_data <- rbind(kiahyundaithefts, kiahyundaidata)
```

# View the combined dataset

```{r head, include=TRUE}
head(combined_data)
```

# --- Visualization 1: Pie Chart ---

```{r pie, include=TRUE}
# Summarize Kia/Hyundai vs Other thefts
theft_counts <- combined_data %>% summarise(Kia_Hyundai 
    = sum(countKiaHyundaiThefts, na.rm = TRUE), Other 
    = sum(countOtherThefts, na.rm = TRUE))

# Reshape data
pie_data <- data.frame( Type = c("Kia/Hyundai", "Other"), 
      Count = c(theft_counts$Kia_Hyundai, theft_counts$Other) )

pie <- ggplot(pie_data, aes(x="", y=Count, fill=Type)) + 
  geom_bar(stat="identity", width=1) + coord_polar("y", start=0) + 
  scale_fill_manual(values = c("Kia/Hyundai" = "red", "Other" = "green")) +
  labs(title="Kia/Hyundai vs Other Vehicle Thefts") + theme_void()

pie
```

# --- Visualization 2: Stacked Bar Chart ---

```{r stack, include=TRUE}
# Summarize thefts by year
yearly_data <- combined_data %>% group_by(year) %>% summarise(Kia_Hyundai = 
  sum(countKiaHyundaiThefts), Other = sum(countOtherThefts)) %>% 
  pivot_longer(cols = c("Kia_Hyundai", "Other"), names_to = "Type", 
  values_to = "Count")

stacked_bar <- ggplot(yearly_data, aes(x=year, y=Count, fill=Type)) + 
  geom_bar(stat="identity") + 
  labs(title="Yearly Kia/Hyundai vs Other Vehicle Thefts", x="Year", 
       y="Number of Thefts")

stacked_bar
```

# --- Visualization 3: Tree Map ---

```{r treemap-plot, fig.width=6, fig.height=4}
# Summarize thefts by city
city_data <- combined_data %>% group_by(city) %>% summarise(Count = 
    sum(countKiaHyundaiThefts)) %>% arrange(desc(Count))

treemap(city_data, index="city", vSize="Count", title="Theft Incidents by City")

```

# --- Visualization 4: Stacked Area ---

```{r Stacked, include=TRUE}
# Summarize yearly thefts
yearly_trends <- combined_data %>% group_by(year) %>% summarise(Kia_Hyundai = 
    sum(countKiaHyundaiThefts, na.rm = TRUE), Other = sum(countOtherThefts, 
    na.rm = TRUE)) %>% pivot_longer(cols = c("Kia_Hyundai", "Other"), names_to =
    "Type", values_to = "Count")

stacked_area_yearly <- ggplot(yearly_trends, aes(x = year, y = Count, fill = 
    Type)) + geom_area() + 
  labs( title = "Trends in Kia/Hyundai vs Other Vehicle Thefts by Year", x = 
          "Year", y = "Number of Thefts" ) + theme_minimal()

stacked_area_yearly
```

# --- Visualization 5: Geospatial Map ---

```{r Geo, include=TRUE}
usa_map <- map_data("state")
geo_map <- ggplot() +
  geom_polygon(data = usa_map, aes(x = long, y = lat, group = group), fill = 
                 "lightgrey", color = "white") +
  geom_point(data = car_map, aes(x = longitude, y = latitude), color 
             = "red", size = 1.25) +
  labs(title = "Geospatial Distribution of Thefts", x = "Longitude", y 
       = "Latitude") +
  theme_void()
geo_map

```

# --- Visualization 6: Heatmap ---

```{r Heat, include=TRUE}
# Monthly thefts data 
month_data <- combined_data %>% group_by(month) %>% summarise( Kia_Hyundai = 
    sum(countKiaHyundaiThefts, na.rm = TRUE), Other = sum(countOtherThefts, 
        na.rm = TRUE) ) %>% pivot_longer(cols = c("Kia_Hyundai", "Other"), 
        names_to = "Type", values_to = "Count") %>% mutate( month = 
        factor(month, levels = month.abb, ordered = TRUE), Type = factor(Type))

# Aggregate data to calculate average percentages by city and year
data_by_year <- combined_data %>% group_by(city, year) %>% 
  summarize(AveragePercentage = mean(percentKiaHyundai, na.rm = TRUE)) %>% 
  ungroup() %>% mutate(year = factor(year, levels = unique(year)), 
  city = factor(city, levels = unique(city)))

ggplot(data_by_year, aes(x = year, y = city, fill = AveragePercentage)) + 
  geom_tile(color = "white") + geom_text(aes(label = 
  scales::percent(AveragePercentage, accuracy = 1)), size = 3) + 
  scale_fill_gradient(low = "#f7f4f9", high = "#49006a", na.value = 
  "grey80", name = "Percentage") + labs(title = 
  "Percentage of stolen vehicles that were Kias or Hyundais (By Year)", 
  x = NULL, y = NULL) + theme_minimal(base_size = 12) + theme(axis.text.x 
  = element_text(angle = 45, hjust = 1, vjust = 1, size = 10), axis.text.y = 
    element_text(size = 10), plot.title = element_text(hjust = 0.5, size = 14), 
  legend.position = "right")
```

# --- Visualization 7: Area Chart ---

```{r Area, include=TRUE}
data_processed <- combined_data %>% group_by(year, city) %>% 
  summarize(AveragePercentage = mean(percentKiaHyundai, na.rm = TRUE)) %>% ungroup()

ggplot(data_processed, aes(x = year, y = AveragePercentage, fill = 
  city, group = city)) + geom_area(alpha = 0.7) + scale_fill_viridis_d(name = "City")+ 
  labs(title = "Yearly Trend of Kia/Hyundai Thefts by City", x = "Year", y 
   = "Average Percentage") + theme_minimal(base_size = 12) + theme(plot.title 
   = element_text(hjust = 0.5, size = 14), legend.position = "bottom", axis.text.x 
   = element_text(angle = 45, hjust = 1, size = 10), axis.text.y = element_text(size = 10))
```
