---
title: "Student Survey Analysis"
author: "Reppeto, Brian"
date: July 18, 2023
output:
  pdf_document: default
  html_document: default
  word_document: default
 
---

# Student Survey Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```

# Questions
```{r, echo=FALSE}
library(readr)
setwd("~/DSC520/Week 7")
student_survey_df <-read.csv("student-survey.csv")
```
## 1.

# Calculating the covariance matrix
<!-- cov_matrix -->
```{r Covariance Matrix, echo=FALSE}
cov_matrix <- cov(student_survey_df[, c("TimeReading", "TimeTV", "Happiness", "Gender")])
cov_matrix

```




<!-- cov_time_read_gender -->

```{r eval=FALSE, include=FALSE}
library(readr)
library(ggplot2)
setwd("~/DSC520/Week 7")
student_survey_df <-read.csv("student-survey.csv")

cov_time_read_gender <-
  cov(student_survey_df$TimeReading, student_survey_df$Gender)
cov_time_read_gender
```

<!-- cov_time_tv_gender -->

```{r eval=FALSE, include=FALSE}
cov_time_tv_gender <- cov(student_survey_df$TimeTV, student_survey_df$Gender)
cov_time_tv_gender
```

<!-- cov_happiness_gender -->

```{r eval=FALSE, include=FALSE}
cov_happiness_gender <-
  cov(student_survey_df$Happiness, student_survey_df$Gender)
cov_happiness_gender

```

The covariance matrix provides valuable insights into the relationships between 
variables. It measures how the variables change together. Based on these results
the covariance for cov_time_read_gender is -0.08181818. This means there is a 
slightly negative relationship between the time spent reading and gender. The 
covariance for cov_time_tv_gender is 0.04545455. This indicates a positive 
relationship between the time spent watching TV and gender.  The covariance for  
cov_happiness_gender is 1.116636, which indicates a relatively strong positive 
relationship between happiness and gender.  All and all, A positive covariance 
indicates a positive relationship, and a negative covariance indicates a 
negative relationship.



## 2.
The Student Survey dataset consists of four variables:
TimeReading: The amount of time spent reading (in minutes).
TimeTV: The amount of time spent watching television (in minutes).
Happiness: Subjective happiness rating.
Gender: Gender of the student (0 for female, 1 for male).

Changing the measurement scale of variables can affect the covariance 
calculation.  For example, if we change the measurement of TimeReading to 
hours instead of minutes, the covariance values would increase by a factor 
of 60 since the unit of measurement is larger. The same could be said about 
changing the measurement units of TimeTV, as it is measured in minutes as well.

A better alternative to address this issue is to standardize the variables 
before calculating the covariance or correlation. Standardizing the variables
scales to have a zero mean and unit variance, would allowing for more 
meaningful comparison.

## 3.
A Pearson correlation test might be a good alternative with this dataset. Since
all the variables are continuous (TimeReading, TimeTV, Happiness), and we are 
interested in measuring the linear relationship between them. Pearson 
correlation is appropriate for this scenario. Based on the Pearson correlation
below, there appears to be a positive correlation between TimeTV and Happiness,
and a negative correlation between TimeReading and TimeTV.

## 4.
Correlation analysis of all variables
```{r Pearson Correlation, echo=FALSE}
cor_matrix_all <-
  cor(student_survey_df [, c("TimeReading", "TimeTV", "Happiness")], method = "pearson")
cor_matrix_all
```



Correlation point graph on Happiness and TimeTv

```{r echo=FALSE}
library(ggplot2)
cor_plot <-ggplot(student_survey_df, aes(x = Happiness, y = TimeTV)) + 
  geom_point() +     geom_smooth(method = "lm",se = TRUE, color = 'red')
cor_plot
```

Single correlation between TimeTv and Happiness

```{r echo=FALSE}
cor_single_pair <-
  cor(student_survey_df$TimeTV,
      student_survey_df$Happiness,
      method = "pearson")
cor_single_pair
```

Confidence correlation with interval at 99%
```{r echo=FALSE}
cor_matrix_all_99 <-
  cor.test(
    student_survey_df$TimeTV ,
    student_survey_df$Happiness,
    method = "pearson",
    conf.level = 0.99
  )
cor_matrix_all_99
```
The correlation matrix suggests that there are significant relationships 
between the variables. Spending more time reading appears to be associated 
with less time spent watching TV and a lower happiness level. Additionally, 
spending more time watching TV is linked to less time spent reading and a 
slightly higher happiness levels.

## 5.
Pearson Correlation coefficient of TimeTV and Happiness 

Coefficient of determination of TimeTV and Happiness
```{r echo=FALSE}
cor_coefficient <-
  cor(student_survey_df$TimeTV,
      student_survey_df$Happiness,
      method = "pearson")
coeff_of_determination <- cor_coefficient ^ 2
cor_coefficient
coeff_of_determination 
```

The results suggest that there is a positive relationship between time and 
happiness, and time can explain about 40.52% of the variability in happiness 
levels. 

## 6.
Based on the analysis, there is a correlations between TimeReading, TimeTV, 
and Happiness. However, remember that correlation does not imply causation.
While I cannot conclude that watching more TV caused students to read less 
based on the analysis. Further more controlled experiments would be necessary 
to establish causation.

## 7.

Partial correlation between TimeReading and Happiness controlling for Gender
```{r echo=FALSE}
library(ppcor)


partial_corr <-
  pcor.test(student_survey_df$TimeReading,
            student_survey_df$Happiness,
            x = student_survey_df$Gender)
print(partial_corr$estimate)
```

The partial correlation between time reading and happiness, controlling for 
gender, indicates a weak negative association. Controlling for the gender 
variable allows us to understand the unique relationship between the other two 
variables and better understand their connection. 
