---
title: "Predicting Fantasy Football Draft Rankings: Unraveling the Enigma"
author: "Brian Reppeto"
date: '2023-08-7'
fontsize:  18pt
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: bibliography.bib
---


```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(lm.beta)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)

```
# Table of Contents

## 1. Introduction

## 2. Approach

## 3. Analysis

## 4. Plots and Visualizations

## 5. Implications

## 6. Concluding Remarks

## 7. Data Collection and Cleaning

## 8. Appendix

## 9. References


# 1. Introduction

Fantasy football is a popular online game where participants act as team 
managers and draft real players from the National Football League (NFL) to form 
fantasy teams. The performance of the actual stats of the players in real-life 
games determines the fantasy team's score. Participants compete against one 
another on a weekly head to head basis and these scores determine a weekly 
winner.  One of the most important aspects of fantasy football is the drafting 
phase, which takes place before the NFL regular season starts.  Participants 
select players based on their projected performance and rankings.  I will use 
the redraft type of league along with a snake draft type of draft.
Additionally, I will use the standard scoring system, which is 1 point per 25 
passing yards, 4 points per passing touchdown, 1 point per 10 rushing or
receiving yards, 6 points per rushing or receiving touchdown, -2 points per
fumble lost or interception.

 
The issue I want to research is being able to help predict fantasy football 
drafting rankings.  Specifically, drafting quarterbacks, running backs, 
wide receivers, and tight ends.  The goal is to develop a model or methodology 
that can help accurately predict the ending rankings of NFL players for fantasy 
football drafts. This project would be of particular interest to fantasy 
football participants, as it can help them better draft a fantasy teams for 
their season.  By leveraging data science techniques, I can analyze historical 
player performance and other relevant data to make informed predictions for the 
upcoming season's draft rankings.  

## Problem statement


Within this vibrant tapestry of fantasy football, a central enigma persists—the 
challenge of accurately predicting the future performance and rankings of NFL 
players in fantasy drafts. The nebulous interplay of player statistics, team 
dynamics, and unforeseen factors complicates this task. Our research aims to 
decipher this puzzle, focusing on forecasting draft rankings for pivotal 
positions: quarterbacks, running backs, wide receivers, and tight ends. 
This endeavor seeks to empower fantasy football enthusiasts by arming them with 
insights grounded in data science, ultimately enhancing their draft strategies 
and bolstering their chances of success in the league.

The core issue revolves around the unpredictability of fantasy football player 
draft rankings.  The rankings and draft shape the core of the team that a player
will play with during the season and understanding how to draft a better team 
can help win a league.  This research will focus on understanding quarterbacks,
running backs, wide receivers, and tight ends as these positions generate the 
majority of points and take up 90 of a fantasy team roster.


# 2. Approach

## Data Collection and Cleaning

The foundation of any data-driven analysis lies in the quality and relevance of 
the data collected. In this phase, I embarked on a meticulous process of data
collection, gathering an extensive dataset that spans multiple seasons 
(2019-2022). The dataset included player statistics, team standings, strength of
schedule (SOS), and offensive team statistics. Each data source was carefully 
selected to provide a comprehensive view of player performance, team dynamics, 
and external factors that could influence fantasy football draft rankings.

1. Data Collection: I gathered an extensive dataset of player statistics from 
past seasons (2019-2022).  I also gathered datasets for strength of schedule
(SOS) and the final standings of the 32 NFL teams from 2022.  The historical 
fantasy data helped tell the story of individual performances during this time.
2. Data Cleaning and Consolidation: The data was cleaned and consolidated, 
ensuring that each player's performance was accurately represented over time. 
This step was critical to ensure the integrity of our analysis and the validity
of our conclusions.  This step of cleaning and consolidating the datasets was
very time consuming as the data has to be standardized and consolidated from 
multiple different sources.
3. Exploratory Analysis: Diving into the data, unveiled trends, patterns, and 
relationships. I investigated the relationship between player performance and 
variables like team SOS (2019-2022) and team standings (2022).
4. Linear Regression Analysis: To assess the predictive potential of the two
variables, I employed a linear regression analysis. I tested the correlation 
between player fantasy points and team SOS (2019- 2022), as well as the 
relationship between fantasy points and team standings (2022). The analyses 
provided insight into the connections between player performance and these
variables.
5. Model Recommendations: While my primary focus was on exploration and analysis, 
I was able to glean insights from my readings that a Random Forest Regression
Model (RFRM) might work best with this set of data.  A RFRM would help with the 
complex relationships between player performance and other variables that might
influence the performance.  Additionally, a RFRM could help with possible 
over fitting and outliers from individual player performances.

## Data

```{r stats, include=FALSE}
library(readxl)
#library(tidyverse)
setwd("~/DSC520/Week 10")
QB_stats<-read_xlsx("QB_Merged_Fantasy_Football_Statistics.xlsx")
RB_stats<-read_xlsx("RB_Merged_Fantasy_Football_Statistics.xlsx")
WR_stats<-read_xlsx("WR_Merged_Fantasy_Football_Statistics.xlsx")
TE_stats<-read_xlsx("TE_Merged_Fantasy_Football_Statistics.xlsx")

```
```{r, include=TRUE}
head(QB_stats)
head(RB_stats)
head(WR_stats)
head(TE_stats)
```
-   footballdb [@scoring2023]

-   footballdb [@fantasydata2023]

```{r SOS, include=FALSE}
library(readxl)
#library(tidyverse)
setwd("~/DSC520/Week 10")
SOS<-read_xlsx("Strength of Schedule.xlsx")

```
```{r SOSall, include=TRUE}
tail(SOS)

```
-   Strength of schedule [@sos2023]

```{r Standings, include=FALSE}
library(readxl)
#library(tidyverse)
setwd("~/DSC520/Week 10")
team_stand_22<-read_xlsx("2022_team_standings.xlsx")

```
```{r, include=TRUE}
head(team_stand_22)
```
-   Team schedule [@NFL2023]

## Sumamry stats
```{r QB_sum, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
QB_summary_data <- QB_stats %>%
  group_by(Player,Posit, Fantasy_yr,Season_team) %>%
  #summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  summarize(FPTS)
  #arrange(desc(Avg_fant_pts))
print(QB_summary_data)
```
```{r RB_summ, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
RB_summary_data <- RB_stats %>%
  group_by(Player,Posit, Fantasy_yr,Season_team) %>%
  #summarize(Avg_fant_pts = round(mean(FPTS))) %>%
   summarize(FPTS)
  #arrange(desc(Avg_fant_pts))
print(RB_summary_data)
```
```{r WR_summ, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
WR_summary_data <- WR_stats %>%
  group_by(Player,Posit , Fantasy_yr, Season_team) %>%
  #summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  #arrange(desc(Avg_fant_pts))
   summarize(FPTS)
print(WR_summary_data)
```

```{r TE_summ, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
TE_summary_data <- TE_stats %>%
  group_by(Player,Posit , Fantasy_yr, Season_team) %>%
  #summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  #arrange(desc(Avg_fant_pts))
   summarize(FPTS)

print(TE_summary_data)
```


```{r consolidated, include=TRUE}
all_players <- rbind(QB_summary_data, RB_summary_data, WR_summary_data, TE_summary_data)#%>%
#filter (Season_team == "ARI")
print(all_players)
```

```{r, grouped points by team & SOS, include=TRUE}
all_sum_team <- all_players %>%
  group_by(Season_team, Fantasy_yr)%>%
  select(Season_team, Fantasy_yr, FPTS)%>%
  summarize(ttl_FPTS = sum(FPTS))%>%
  arrange(desc(Season_team))%>%
  filter (Season_team != "NA")
  
  

print(all_sum_team)

```
```{r avg_sos_rank, include=TRUE}
sos_rank <- SOS %>%
  group_by(Team, Year)%>%
  select(Team, Year, Last)%>%
  arrange(Team,Year)%>%
  filter (Team != "NO TEAM FA")
  #summarize(avg_sos  = round(mean(ranking)))
# the higher the ranking the harder the schedule i.e 1 is the hardest

print(sos_rank)
```
```{r, include=TRUE}
library(dplyr)


all_sum_team <- all_sum_team %>%
  left_join(sos_rank, by = c("Season_team" = "Team", "Fantasy_yr" = "Year"))
  #select(-ranking, -Last.x)  # Removing duplicate columns

print(all_sum_team)
```







# 3. Analysis


## Performance Patterns Across Positions

The initial phase of analysis revolved around understanding the performance 
patterns across different player positions. The heatmap of positional scores 
provided a visual representation of the average fantasy points earned by 
quarterbacks (QBs), running backs (RBs), wide receivers (WRs), and tight ends 
(TEs) over the seasons 2019-2022. This visualization revealed distinct patterns 
in point distribution among these positions. Notably, it underscored the 
scarcity of high-scoring RBs, indicating a potential draft strategy of 
prioritizing top-performing RBs early due to their scarcity.

Influence of Team Success and Strength of Schedule

One of the key insights we uncovered was the profound influence of team dynamics
on player performance. By examining the relationship between team success and 
player fantasy points, we observed that players belonging to successful teams 
tended to generate higher fantasy point outputs. This phenomenon highlights the 
interplay between team success and individual player performance, indicating 
that drafting players from teams with higher winning percentages could be a 
strategic advantage.

However, the relationship between player performance and strength of schedule 
(SOS) was more nuanced. While there was a degree of correlation between SOS and 
fantasy points, the predictive power of SOS alone was limited. This indicated 
that the difficulty of opponents, as represented by SOS, is not a definitive 
predictor of player performance. Fantasy managers should consider other 
variables, such as team success, when making drafting decisions.

Insights from Linear Regression Analysis

We conducted linear regression analyses to quantify the relationships between 
fantasy points, team success, and SOS. The regression model for team standings 
exhibited a substantial R-squared value, indicating that team success 
(measured by winning percentage) is a significant predictor of player 
performance. Fantasy football managers should take note of this relationship, 
as players from successful teams are more likely to deliver higher fantasy 
point outputs.

In contrast, the regression model for SOS demonstrated a weaker R-squared value,
implying that SOS alone has limited predictive power for fantasy points. This 
underscores the complexity of player performance, suggesting that factors beyond
SOS contribute significantly to a player's fantasy point output.

Additionally, the heat map of historical player performance provides a roadmap 
for which direction related to player position when drafting in a snake draft.
There were 35 QB's whom had more than 300 points since 2019 (per year), there 
were 3 RB's whom had more than 300 points since 2019 (per year), there were 14 
WR's who had more than 200 points since 2019 (per year), and only 6 TE's whom
had more than 150 points (per year).  These facts point out that the shortage
of a good RB would create a higher demand and should be drafted earlier than all
other positions assuming it is one of the top 3.  Next import would be one of 
the top 5 QB's whom were over 400 points (per year), or the top WR whom was just 
shy of 300 point (per year).  Once these players are gone, the distinction 
of position to draft next becomes a little less clear, as there are only a 
handful of second tier RB's to take with just two hovering around the 300
mark. 


```{r lm SOS, include=TRUE}
library(lm.beta)
model_1 <- lm( ttl_FPTS ~ Last, data = all_sum_team)
summary(model_1)
standardized_betas_model_1 <- lm.beta(model_1)
#all_sum_team$std_beta_sos_rank <- standardized_betas_model_1[2]
print(standardized_betas_model_1)
#print(all_sum_team)

conf_intervals_model_1 <- confint(model_1)



print(conf_intervals_model_1)


```

The R-squared values are low, but suggest that the model can explain only a 
small portion of variance.  While there is a relationship between the SOS and 
Fantasy point, it is probably not a good indicator of future fantasy points.


```{r 22 team, include=TRUE}
all_22_team <- all_players %>%
  filter (Fantasy_yr == "2022", !is.na(Season_team))%>%
  group_by(Season_team) %>%
  summarise(total_FPTS = sum(FPTS))%>%
  left_join(team_stand_22, by = c("Season_team" = "Team"))
 


  print(all_22_team)

```


```{r lm Standings, include=TRUE}
library(lm.beta)
model_2 <- lm( total_FPTS ~ Pct, data = all_22_team)
summary(model_2)
standardized_betas_model_2 <- lm.beta(model_2)
#all_sum_team$std_beta_sos_rank <- standardized_betas_model_1[2]
print(standardized_betas_model_2)
#print(all_sum_team)

conf_intervals_model_2 <- confint(model_2)



print(conf_intervals_model_2)


```
Overall, the winning percentage variable indicates that the "Pct" variable has 
a significant impact on predicting the "total_FPTS," and the model explains a 
substantial portion of the variability in the dependent variable.  This variable
will be good to help predict the more valuable players.  If your team has a 
higher winning percentage the player will be more valuable as they will score 
more points.


# 4. Plots and Visualizations

## Plots and tables needed

1. Line plots to visualize trends in player performance over time.
2. Tables to display the top-performing players based on historical fantasy 
points.
3. Heat map to help understand which positions are more important to draft due 
to outliers in scoring by position.




```{r QB_by_year ,echo=FALSE}
library(dplyr)
library(ggplot2)
# Create the heatmap using ggplot2
heatmap <-
  ggplot(all_players, aes(x = Posit, y = FPTS, fill = FPTS)) +
  geom_tile(color = "black ") +
  scale_fill_gradient(low = "orange", high = "blue") +
  theme_minimal() +
  labs(x = "Position", y = "Fantasy Points", 
       title = "Heatmap of Scores by Player Avg 2019-2022")


print(heatmap)

```


The heat map shows some very important features when it comes to drafting 
analysis. There are a significant number of Qb's with more than 300 points, and 
only a small handful of Rb's that reach this milestone.  



```{r, inlcude=FALSE}
QB_filtered_players <-
  subset(QB_summary_data, Season_team != "NO TEAM FA" &
           FPTS >= 100)
RB_filtered_players <- subset(RB_summary_data, Season_team != "NO TEAM FA")
WR_filtered_players <- subset(WR_summary_data, Season_team != "NO TEAM FA")
TE_filtered_players <- subset(TE_summary_data, Season_team != "NO TEAM FA")
```


```{r line_plot, echo=FALSE}
library(ggplot2)

ggplot(QB_filtered_players , aes(x = Player, y = FPTS, group = .5)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Fantasy Points QB's >=100 pts",
       x = "Player Name",
       y = "Average Fantasy Points") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size =5))
        plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")

```




# 5. Implications

The analysis of fantasy football data, provided several intriguing insights 
on the intricate dynamics of player performance and their 
relationship with team attributes:

Team Success and Player Performance: A compelling connection was unearthed 
between team success and player performance. Players belonging to teams with 
higher winning percentages tended to produce higher fantasy points. This 
revelation challenges the notion that player performance exists in isolation and
underscores the profound influence of team dynamics on individual output.
Limited Predictive Power of SOS: While the strength of schedule (SOS) metric 
demonstrated a correlation with player fantasy points, its predictive power
proved to be limited. The notion that a harder or easier schedule directly 
translates to better or worse player performance was debunked. This insight 
prompts a reconsideration of the common belief in SOS as a definitive predictor.
Positional Differences: The heat map analysis illustrated the stark contrast in 
fantasy points distribution across positions. Quarterbacks boasted a wider range
of fantasy points, with numerous players crossing the 300-point threshold. 
Conversely, running backs exhibited fewer high-scoring players, suggesting a 
position-specific variability in performance.
Draft Strategy Implications: The revelation of team dynamics' impact on player 
performance introduces a strategic edge in drafting. Managers who prioritize 
players from successful teams might enhance their fantasy point accumulation. 
This insight empowers managers to strategize beyond individual player statistics
and delve into team statistics for a more holistic drafting approach.
Foundation for Future Predictive Models: While I didn't construct an extensive 
predictive model in this phase, the relationships uncovered lay the groundwork 
for future modeling endeavors. The insights guide the selection of features and 
the creation of models that can better predict fantasy football draft rankings.
Complexity of Player Performance: The analysis highlighted the multifaceted 
nature of player performance prediction. Player statistics alone cannot 
encapsulate the intricate web of factors affecting fantasy output. Team 
dynamics, SOS, and other external variables collectively shape player 
performance. These insights collectively redefine the narrative of fantasy 
football drafting, emphasizing the significance of team success, the nuanced 
interplay between variables, and the need for predictive models that capture 
these complexities. As enthusiasts navigate the realm of fantasy football, armed
with this new found understanding, they're poised to make more informed 
decisions, thereby elevating their chances of championship glory.



# 6. Concluding remarks

While the analysis provides valuable insights, it's essential to recognize its 
limitations and opportunities for improvement:

Limited Sample Size: The analysis spanned a few years of historical 
data (2019-2022).  This short period of analysis might not encompass all 
possible scenarios and trends. Expanding the data set to include more seasons 
could provide a broader perspective on player performance.
Model Development: While the analysis explored relationships between variables, 
I didn't construct a predictive model. Developing and fine-tuning a predictive 
model or models could provide more insights for fantasy football managers.
Advanced statistics: My analysis mainly focused on existing features an did not 
incorporating any additional features, such as player performance ratios or 
advanced statistics.
External Factors: I didn't delve into external factors that could influence 
player performance, such as injuries, player trades, coaching changes or weather. 
Model Selection: While Random Forest Regression is a strong contender, other 
advanced models like Gradient Boosting, Neural Networks, or time series models 
could be explored. A model comparison study could help identify the most 
suitable model for this specific problem.
Real-time Updates: The analysis used historical data, limiting its real-time 
relevance. Creating a system that updates predictions during the NFL season 
based on current player performance and team dynamics could provide actionable 
insights for fantasy managers.

In summary, while the analysis has provided several key insights, its potential 
can be unlocked by addressing these limitations. Future iterations could 
integrate advanced modeling techniques, more comprehensive datasets, and a 
multidimensional approach that considers player statistics, and external 
factors. By addressing these challenges, the analysis could evolve from an 
exploratory exercise into a powerful tool guiding fantasy football managers 
towards successful drafting strategies.


# 7. Data collection and cleaning


## Datsets

Strength of schedule (as of 9/1 each year):

https://www.teamrankings.com/nfl/ranking/schedule-strength-by-other?date=2022-09-01

Fantasy scoring points by two types of systems:

https://www.footballdb.com/fantasy-football/index.html

Team standings:
https://www.nfl.com/standings/division/2022/REG

Player seasonal teams:
https://fantasydata.com/nfl/fantasy-football-leaders

## Required Packages

'tidyverse'
'ggplot2' 
'dplyr'
'lm.beta'





### Fantasy stats
The data had to have positions added to the data field, in addition to the 
fantasy year. Additionally, each position file had to be downloaded separately 
and combined into one one fantasy stats table per position for 2019-2022. 
Moreover, the fields also had to have the team for each player added.  Since the
fantasy stats reflected the players current team, these had to added to the file
by inserting them into each of the roles.


### Team standings
The data had to be put into a tibble format removing the division and conference.
Additionally, the teams moving on to the playoffs had to have the signifying 
symbols removed from the team name.  Moreover, the fields also were 
standardized by unbolding, removing the underlines, and hyperlinks.  The team
names had to be standardized to match the abbreviation used in the fantasy 
numbers files.


### Team offensive stats
The offensive team stats had to have the fields standardized by unbolding, 
removing the underlines, and hyperlinks.


### Strength of schedule (SOS)
The SOS data set had to have the date added to the file.  Each file was 
downloaded separately and appended to each file.  The file is as of 2/28 for the 
year that the season started (2019-2022).  This allows for the final SOS to have
been calculated.  The team names had to be standardized to match the 
abbreviation used in the fantasy 
numbers files. 




# 8. Appendix


## Cleaning data
```{r QB, eval=FALSE, include=TRUE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise
data1 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB-2_2021.csv")
data2 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB_2022.csv")
data3 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB-3_2020.csv")
data4 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB-4_2019.csv")

# Combine all data frames into one
combined_data <-
  rbind(data1,data2,data3,data4)

# Write the merged data frame to a new CSV file
write.csv(combined_data, file = "QB_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)


```
```{r RB, eval=FALSE, include=TRUE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise

data5 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB-2_2021.csv")
data6 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB-3_2020.csv")
data7 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB-4_2019.csv")
data8 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB_2022.csv")


# Combine all data frames into one
combined_data2 <-
  rbind(data5,data6,data7,data8)

# Write the merged data frame to a new CSV file
write.csv(combined_data2, file = "RB_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)


```
```{r WR, eval=FALSE, include=TRUE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise

data9 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR-2_2021.csv")
data10 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR-3_2020.csv")
data11 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR_2022.csv")
data12 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR-4_2019.csv")

# Combine all data frames into one
combined_data3 <-
  rbind(data9,data10,data11,data12)

# Write the merged data frame to a new CSV file
write.csv(combined_data3, file = "WR_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)

```
```{r TE, eval=FALSE, include=TRUE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise

data13 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE-2_2021.csv")
data14 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE-4_2019.csv")
data15 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE_2022.csv")
data16 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE-3_2020.csv")

# Combine all data frames into one
combined_data4 <-
  rbind(data13,data14,data15,data16)

# Write the merged data frame to a new CSV file
write.csv(combined_data4, file = "TE_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)


```

# 9. References
```{r bibliography, echo=FALSE}

```













