---
title: "Final Step 2"
author: "Brian Reppeto"
date: '2023-08-6'
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: bibliography.bib
---

# Markdown Basics

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Introduction

Fantasy football is a popular online game where participants act as team 
managers and draft real players from the National Football League (NFL) to form 
fantasy teams. The performance of the actual stats of the players in real-life 
games determines the fantasy team's score. Participants compete against one 
another on a weekly head to head basis and these scores determine a weekly 
winner.  One of the most important aspects of fantasy football is the drafting 
phase, which takes place before the NFL regular season starts.  Participants 
select players based on their projected performance and rankings.  I will use 
the redraft type of league along with a snake draft type of draft for this
analysis. Additionally, I will use the standard scoring system, which consists 
of the following score process, 1 point per 25 passing yards, 4 points per 
passing touchdown, 1 point per 10 rushing or receiving yards, 6 points per 
rushing or receiving touchdown, -2 points per fumble lost or interception.


The issue I want to research is being able to help predict fantasy football 
drafting rankings and determine the value of positions in the draft.  
Specifically, drafting quarterbacks, running backs, wide receivers, and tight 
ends.  The goal is to develop a model or methodology that can help accurately 
predict the ending rankings of NFL players for fantasy football drafts. This 
project would be of particular interest to fantasy football participants, as it 
can help them better draft a fantasy teams for their season.  By leveraging data
science techniques, I can analyze historical player performance and other 
relevant data to make informed predictions for the upcoming season's draft 
rankings.  




## Research questions

1. What statistical features of NFL players historical performances are most 
indicative of their fantasy football draft rankings?
2. Can we identify any patterns or trends in player performance over the past 
few seasons that might help predict future draft rankings (rookies will be 
ignored for this evaluation)?
3. Is there a relationship between the SOS of a team and the points scored by 
the players for its team?
4. Are there specific positions (quarterback, running back, wide receiver, etc.)
that are more valuable to draft in a fantasy league before others? 
5. Is there a relationship between an NFL teams winning percentage and the
fantasy points scored by its players?


## Approach

1. Data Collection: Gather historical player performance data, fantasy football 
draft rankings data, team data, and any other relevant datasets. 
2. Conduct data analysis to understand the data, identify any missing values or 
outliers, and gain insights into player performance and rankings trends.
3. Select or create relevant features that may impact draft rankings (i.e. 
player statistics from previous seasons, team performance metrics (Strength of
schedule), etc.)
4. Create a heat map to show the most valuable selections for a snake type
draft.
5. Provide a recommendation for the most effective method to help predict 
fantasy football drafting rankings.


## How approach addressed the problem

By combining the multiple approaches listed above I will be able to compare 
individual performances along with other factors to help provide valuable 
insights into a ranking and value list for fantasy football managers.  This will
all them to build competitive teams for the fantasy season. This will also allow
me to group the players into clusters.

This project will not be fully operational as all cuts for the NFL season will 
not have happened by the time this project is complete.


## Cleaning data
```{r QB, eval=FALSE, include=FALSE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise
data1 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB-2_2021.csv")
data2 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB_2022.csv")
data3 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB-3_2020.csv")
data4 <- read.csv("FantasyPros_Fantasy_Football_Statistics_QB-4_2019.csv")

# Combine all data frames into one
combined_data <-
  rbind(data1,data2,data3,data4)

# Write the merged data frame to a new CSV file
write.csv(combined_data, file = "QB_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)


```
```{r RB, eval=FALSE, include=FALSE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise

data5 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB-2_2021.csv")
data6 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB-3_2020.csv")
data7 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB-4_2019.csv")
data8 <- read.csv("FantasyPros_Fantasy_Football_Statistics_RB_2022.csv")


# Combine all data frames into one
combined_data2 <-
  rbind(data5,data6,data7,data8)

# Write the merged data frame to a new CSV file
write.csv(combined_data2, file = "RB_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)


```
```{r WR, eval=FALSE, include=FALSE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise

data9 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR-2_2021.csv")
data10 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR-3_2020.csv")
data11 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR_2022.csv")
data12 <- read.csv("FantasyPros_Fantasy_Football_Statistics_WR-4_2019.csv")

# Combine all data frames into one
combined_data3 <-
  rbind(data9,data10,data11,data12)

# Write the merged data frame to a new CSV file
write.csv(combined_data3, file = "WR_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)

```
```{r TE, eval=FALSE, include=FALSE}
# Load the required library
library(tidyverse)

# Step 1: Read the CSV file into a data frame
setwd("~/DSC520/Week 9/csv")
# Read all the CSV files and bind them row-wise

data13 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE-2_2021.csv")
data14 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE-4_2019.csv")
data15 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE_2022.csv")
data16 <- read.csv("FantasyPros_Fantasy_Football_Statistics_TE-3_2020.csv")

# Combine all data frames into one
combined_data4 <-
  rbind(data13,data14,data15,data16)

# Write the merged data frame to a new CSV file
write.csv(combined_data4, file = "TE_Merged_Fantasy_Football_Statistics.csv"
          , row.names = FALSE)


```


### Fantasy stats
The data had to have positions added to the data field, in addition to the 
fantasy year. Additionally, each position file had to be downloaded separately 
and combined into one one fantasy stats table per position for 2019-2022. 
Moreover, the fields also had to have the team for each player added.


### Team standings
The data had to be put into a tibble format removing the division and conference.
Additionally, the teams moving on to the playoffs had to have the signifying 
symbols removed from the team name.  Moreover, the fields also were 
standardized by unbolding, removing the underlines, and hyperlinks.


### Team offensive stats
The offensive team stats had to have the fields standardized by unbolding, 
removing the underlines, and hyperlinks.


### Strength of schedule (SOS)
The SOS data set had to have the date added to the file.  Each file was 
downloaded separately and appended to each file.  The file is as of 2/28 for the 
year that the season started (2019-2022).  This allows for the final SOS to have
been calculated.


## Data

```{r stats, include=FALSE}
library(readxl)
library(tidyverse)
setwd("~/DSC520/Week 9")
QB_stats<-read_xlsx("QB_Merged_Fantasy_Football_Statistics.xlsx")
RB_stats<-read_xlsx("RB_Merged_Fantasy_Football_Statistics.xlsx")
WR_stats<-read_xlsx("WR_Merged_Fantasy_Football_Statistics.xlsx")
TE_stats<-read_xlsx("TE_Merged_Fantasy_Football_Statistics.xlsx")

```
```{r, include=TRUE}
head(QB_stats)
head(RB_stats)
head(WR_stats)
head(TE_stats)
```
-   footballdb [@scoring2023]

```{r SOS, include=FALSE}
library(readxl)
library(tidyverse)
setwd("~/DSC520/Week 9")
SOS<-read_xlsx("Strength of Schedule.xlsx")

```
```{r SOSall, include=TRUE}
tail(SOS)

```
-   Strength of schedule [@sos2023]

```{r Standings, include=FALSE}
library(readxl)
library(tidyverse)
setwd("~/DSC520/Week 9")
team_stand_22<-read_xlsx("2022_team_standings.xlsx")

```
```{r, include=TRUE}
head(team_stand_22)
```

## Sumamry stats
```{r QB_sum, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
QB_summary_data <- QB_stats %>%
  group_by(Player,Posit, Current_team) %>%
  summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  arrange(desc(Avg_fant_pts))
print(QB_summary_data)
```
```{r RB_summ, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
RB_summary_data <- RB_stats %>%
  group_by(Player,Posit, Current_team) %>%
  summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  arrange(desc(Avg_fant_pts))
print(RB_summary_data)
```
```{r WR_summ, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
WR_summary_data <- WR_stats %>%
  group_by(Player,Posit, Current_team) %>%
  summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  arrange(desc(Avg_fant_pts))
print(WR_summary_data)
```

```{r TE_summ, include=TRUE} 
library(dplyr)
# Group data by Player and calculate summary statistics
TE_summary_data <- TE_stats %>%
  group_by(Player,Posit, Current_team) %>%
  summarize(Avg_fant_pts = round(mean(FPTS))) %>%
  arrange(desc(Avg_fant_pts))

print(TE_summary_data)
```


```{r consolidated, include=TRUE}
all_players <- rbind(QB_summary_data, RB_summary_data, WR_summary_data, TE_summary_data)

print(all_players)
```

```{r, grouped points by team & SOS, include=TRUE}
all_sum_team <- all_players %>%
  group_by(Current_team) %>%
  summarize(ttl_Avg_fant_pts =sum(Avg_fant_pts))

print(all_sum_team)

```
```{r avg_sos_rank, include=TRUE}
sos_avg_rank <- SOS %>%
  group_by(Team) %>%
  summarize(avg_sos  = round(mean(ranking)))
# the higher the ranking the harder the schedule

print(sos_avg_rank)
```
```{r, include=TRUE}

all_sum_team <- all_sum_team %>% mutate(avg_sos_rank = sos_avg_rank$avg_sos)

print(all_sum_team)
```
```{r lm SOS, include=TRUE}
library(lm.beta)
model_1 <- lm( ttl_Avg_fant_pts ~ avg_sos_rank, data = all_sum_team)
summary(model_1)
standardized_betas_model_1 <- lm.beta(model_1)
print(standardized_betas_model_1)


```




## Datsets

Strength of schedule (as of 9/1 each year):

https://www.teamrankings.com/nfl/ranking/schedule-strength-by-other?date=2022-09-01

Fantasy scoring points by two types of systems:

https://www.footballdb.com/fantasy-football/index.html

Team standings:
https://www.nfl.com/standings/division/2022/REG

## Required Packages
'tidyverse'
'ggplot2' 
'dplyr'
'lm.beta'

## Plots and tables needed

1. Line plots to visualize trends in player performance over time.
2. Tables to display the top-performing players based on historical fantasy 
points.
3. Heat map to help understand which positions are more important to draft due 
to outliers in scoring by position.


## Plots

```{r QB_by_year ,echo=FALSE}
library(dplyr)
library(ggplot2)
# Create the heatmap using ggplot2
heatmap <-
  ggplot(all_players, aes(x = Posit, y = Avg_fant_pts, fill = Avg_fant_pts)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(x = "Position", y = "Average Fantasy Points", 
       title = "Heatmap of Scores by Player Avg 2019-2022")


print(heatmap)

```
```{r, inlcude=FALSE}
QB_filtered_players <-
  subset(QB_summary_data, Current_team != "NO TEAM FA" &
           Avg_fant_pts >= 100)
RB_filtered_players <- subset(RB_summary_data, Current_team != "NO TEAM FA")
WR_filtered_players <- subset(WR_summary_data, Current_team != "NO TEAM FA")
TE_filtered_players <- subset(TE_summary_data, Current_team != "NO TEAM FA")
```


```{r line_plot, echo=FALSE}
library(ggplot2)

ggplot(QB_filtered_players , aes(x = Player, y = Avg_fant_pts, group = .5)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Fantasy Points QB's >=100 pts",
       x = "Player Name",
       y = "Average Fantasy Points") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size =5))
        plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")

```


## Question for future steps

1. How can I incorporate additional data, such as team dynamics (coaching 
issues, player issues), weather conditions, or player off-field behavior, 
to improve the model's predictions?
2. Can I build an ensemble model that combines the strengths of multiple 
algorithms to achieve better prediction accuracy?
3. How can I continuously update the model during the NFL season to account 
for emerging trends and player performance changes?
4. Can I use this research for in-season predictors?
5. Review the correlations and adjust for outliers, also adjust for historical
teams.
6. Once historical teams added to player then review correlation for winning
percentages to player FFB points totals.
7. Provide a recommendation for the most effective method to help predict 
fantasy football drafting rankings.

# References
```{r bibliography, echo=FALSE}

```














